Homeostatic mechanisms may shape the type and duration of oscillatory modulation.

### 2020-05-04 main.tex
Reviewer response 2 - complete

Specifically, we study Ca\textsuperscript{2+}-mediated homeostasis in a feed-forward population of 100 Hodgkin-Huxley neurons. By feedforward we mean neurons in the population are driven only by the inputs to the model. Cells have no lateral or recurrent connections. We modulate this population with neural oscillations. Oscillatory input for each cell in Hodgkin-Huxley population was modelled as a set of about $M=100$ Poisson neurons (Figure~\ref{fig:f1}\textbf{b}-\textbf{c}). With a probability of 0.1 these were randomly drawn from a pool of 1000 Poisson neurons, each of which was independently simulated. To change the level of oscillatory modulation strength, we alter the firing rate of the entire Poisson population. Average synaptic strengths were fixed, but drawn from a uniform distribution for each simulation (see \textit{Methods}). An oscillation's temporal duration was simulated as either tonic (Figure~\ref{fig:f1}\textbf{b}, top), which lasted for the entire experiment (20 sec), or as a 3-cycle burst (Figure~\ref{fig:f1}\textbf{b}, bottom). Bursts were delivered only at the end of a trial, where they overlap with the input stimulus (Figure~\ref{fig:f1}\textbf{a}-\textbf{b}). The stimulus input was also taken from an independent set of about $M=100$ Poisson cells. Stimulus and oscillation populations were independently simulated. The stimulus lasted 0.5 seconds, and had an average firing rate of 6 Hz, and was fixed for all experiments. During the stimulus window we measured the population's firing rate and synchrony. It is these numbers we report throughout.

@@

The general effect we would expect for some other selection of channels, chosen by some other theorists, or by evolution itself, depends strongly on those selected channels. O'Leary \cite{OLeary2014} has made it clear how complex the relationship is between homeostasis, channel conductance, and firing patterns. It's so complex that we believe general statements are probably impossible, at least as they might relate to this paper. While our simple model has been shown by others to recapitulate hippocampal pyramidal cell's activity in laboratory settings \cite{LeMasson1993}, it is a \textit{very simple} model, and validating our work does require new experimental testing. This could be done, for example, by using prolonged excitatory gamma oscillations--arising from direct stimulation of either the dentate gyrus or select areas in CA3 \cite{Csicsvari2003}--to drive a possible homeostatic response in CA1.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{fig3.png}
\caption{\label{fig:f3}
    The effect of fast AMPAergic gamma oscillations for both tonic oscillation and 3-cycle bursts. 
    Top panel (\textbf{a}.) is the observed population firing rate averaged over a 0.05 second or 0.033 second stimulus. Bottom (\textbf{b}.) is synchrony over the same period. Columns in both \textit{a.} and \textit{b.} represent with 60 or 90 Hz gamma. The stimulus window was matched to the length of time 3 cycles of oscillation would take.
}
\end{figure}

@@

We chose a 3-cycle length for bursts somewhat arbitrarily. Real oscillations range in duration from single cycles to seconds of even minutes \cite{Lundqvist2016,vanEde2018}. Our burst/tonic distinction is therefore a false (but useful) dichotomy. To find a better estimate of how long an oscillation would need to last, on average, to generate a homeostatic response we ran some long simulations using a more realistic homeostatic time constant ($\tau_h$ = 600 seconds; up from 20 seconds). It seems an oscillation would need to be present \textit{at least} 10\% of the time to generate a response (Figure~\ref{fig:f5}\textbf{c}). 
Real theta rhythms drift in frequency \cite{Buzsaki2015}. To test if small differences in frequency have a notable effect, we simulated small drifts in frequency of theta ranging from 6-10 Hz. Frequency drift had no visible effect on the homeostatic suppression of excitability (Figure~\ref{fig:f5}\textbf{d}-\textbf{e}).

@@

Our simple models offer answers to three basic questions. One, do tonic oscillations engage homeostatic mechanisms? Two, does homeostasis in turn change the oscillation's function? Three, do short bursts of oscillation have distinct effects from tonic oscillations? Put another way: can homeostasis explain why some oscillations, such as hippocampal theta, tend to appear as tonic rhythms while other oscillations (like those in cortex) tend to appear as bursts?

Oscillations are generally conceived to play a functional role in modulating spiking {Cite Varela}{Cite Fries}, by grouping action potentials and improving signal to noise \cite{Chen2013,Zhou2015,Voytek2015a,Peterson2017}. Absent the effects of homeostasis, our model clearly demonstrates this: with increasing modulation strength, excitatory oscillations show increased firing rate and synchrony. Given this effect, and the importance of phase coding in the hippocampus {Cite}, it is reasonable to wonder why hippocampal oscillations are GABAergic. Our model suggests that while this is true for brief, bursty excitatory oscillations, stronger, tonic excitatory oscillations engage homeostatic mechanisms that alter Na and K conductances, driving down the membrane voltage. This, in turn, reduces both the firing rate and spike synchrony in response to inputs.

@@

Bursting, rather than sustained, oscillations tend to be common in the cortex. One striking example is in motor cortical regions, where beta (~12-30 Hz) bursts are prevalent and likely functional. Specifically, beta bursts are very short--sometimes lasting only one or two cycles \cite{Sherman2016}--and relate to self-timed movements \cite{Feingold2015}. Patients with Parkinson's disease show increased motor rigidity and bradykinesia, symptoms associated with prolongation of beta bursts \cite{Tinkhauser2017}. Levodopa treatment was shown to decrease burst probability and duration, and that decrease in burst duration correlated with motor improvement \cite{Tinkhauser2017}.

An important counterexample to beta is visual cortical alpha (8-10 Hz). Alpha is tonic and high powered when humans rest with their eyes closed. Even though both cortical and sub-cortical alpha generators are synaptically excitatory, this rhythm has however been associated with suppression of excitability \cite{Jensen2002,Bonnefond2012,Peterson2017}. While several competing explanations have been offered for this \cite{Bonnefond2012,Lange2013,Peterson2017}, our work raises another possibility that is complementary to the others. Specifically, though we modelled hippocampal cells, the same Ca\textsuperscript{2+} homeostatic mechanisms exist in visual cortex (and the neocortex as a whole). This means that strong and tonic alpha oscillations, combined with homeostasis, should directly suppress population firing in visual cortex. Such an effect, were it to occur, would last well past the moment of oscillation offset. In fact, such long term effects of alpha have been reported in the literature \cite{Jensen2002,Bonnefond2012}, though the physiological mechanism was often unclear. Our work suggests that intrinsic homeostasis may underlie this effect.

@@

% TODO add noise term here. Replaces r_b. You have not mentioned this until now why you idiot?

>>>>

Specifically, we study Ca\textsuperscript{2+}-mediated homeostasis in a feed-forward population of 100 Hodgkin-Huxley neurons. By feedforward we mean neurons in the population are driven only by the inputs to the model. Cells have no lateral or recurrent connections. We modulate this population with neural oscillations. Oscillatory input for each cell in Hodgkin-Huxley population was modeled as a set of about $M=100$ Poisson neurons (Figure~\ref{fig:f1}\textbf{b}-\textbf{c}). With a probability of 0.1 these were randomly drawn from a pool of 1000 Poisson neurons, each of which was independently simulated. To change the level of oscillatory modulation strength, we altered the firing rate of the entire Poisson population. Average synaptic strengths were fixed, but drawn from a uniform distribution for each s

imulation (see \textit{Methods}). An oscillation's temporal duration was simulated as either tonic (Figure~\ref{fig:f1}\textbf{b}, top), which lasted for the entire experiment (20 sec), or as a 3-cycle burst (Figure~\ref{fig:f1}\textbf{b}, bottom). Bursts were delivered only at the end of a trial, where they overlap with the input stimulus (Figure~\ref{fig:f1}\textbf{a}-\textbf{b}). The stimulus input was also taken from an independent set of about $M=100$ Poisson cells. Stimulus and oscillation populations were independently simulated. The stimulus lasted 0.5 seconds, and had an average firing rate of 6 Hz, and was fixed for all experiments. During the stimulus window we measured the population's firing rate and synchrony. It is these numbers we report throughout.

@@

The general effect we would expect for some other selection of channels, chosen by some other theorists, or by evolution itself, depends strongly on those selected channels. O'Leary \cite{OLeary2014} has made it clear how complex the relationship is between homeostasis, channel conductance, and firing patterns. It's so complex that we believe general statements are probably impossible, at least as they might relate to this paper. While our simple model has been shown by others to recapitulate hippocampal pyramidal cell's activity in laboratory settings \cite{LeMasson1993}, it is a \textit{very simple} model, and validating our work does require new experimental testing. This could be done, for example, by using prolonged excitatory gamma oscillations--arising from direct stimulation of either the medial entorhinal cortex or CA3--to drive a possible homeostatic response in CA1 \cite{Csicsvari2003,Colgin2009}.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{fig3.png}
\caption{\label{fig:f3}
    The effect of fast AMPAergic gamma oscillations for both tonic oscillation and 3-cycle bursts. 
    Top panel (\textbf{a}.) is the observed population firing rate averaged over a 0.05 second or 0.033 second stimulus. Bottom (\textbf{b}.) is synchrony over the same period. Columns in both \textit{a.} and \textit{b.} represent with 60 or 90 Hz gamma. The stimulus window was matched to the length of time 3 cycles of oscillation would take. To generate this curve the model was equilibrated for 20 seconds, under a 1.5 Hz firing rate and 8 Hz frequency rhythm, after which the current impulse that we described above was applied. During FI-curve estimation these was no oscillation present.
}
\end{figure}

@@

We chose a 3-cycle length for bursts somewhat arbitrarily. Real oscillations range in duration from single cycles to seconds of even minutes \cite{Lundqvist2016,vanEde2018}. Our burst/tonic distinction is therefore a false (but useful) dichotomy. To find a better estimate of how long an oscillation would need to last, on average, to generate a homeostatic response we ran some long simulations using a more realistic homeostatic time constant ($\tau_h$ = 600 seconds; up from 4 seconds). It seems an oscillation would need to be present \textit{at least} 10\% of the time to generate a response (Figure~\ref{fig:f5}\textbf{c}). Jacobs \cite{Jacobs2014} estimated the prevalence of hippocampal theta to be about 40\% of the time in rats, and 10\% in humans. So even for humans, who have slower and less stable theta rhythm(s) \cite{Goyal2018}, we'd expect an homeostatic response from AMPAergic theta, if it existed.

Real theta rhythms drift in frequency \cite{Buzsaki2015,Jacobs2014}. To test if small differences in frequency have a notable effect, we simulated small drifts in frequency of theta ranging from 6-10 Hz. Frequency drift had no visible effect on the homeostatic suppression of excitability (Figure~\ref{fig:f5}\textbf{d}-\textbf{e}).

@@

Our simple models offer answers to three basic questions. One, do tonic oscillations engage homeostatic mechanisms? Two, does homeostasis in turn change the oscillation's function? Three, do short bursts of oscillation have distinct effects from tonic oscillations? Put another way: can homeostasis explain why some oscillations, such as hippocampal theta, tend to appear as tonic rhythms while other oscillations (like those in cortex) tend to appear as bursts? \cite{Peterson2017,Cole2018,vanEde2018a}

Oscillations are generally conceived to play a functional role in modulating spiking \cite{Fries2015a,vanEde2018b} by grouping action potentials and improving signal to noise \cite{Chen2013,Zhou2015,Voytek2015a,Peterson2017}. Absent the effects of homeostasis, our model clearly demonstrates this: with increasing modulation strength, excitatory oscillations show increased firing rate and synchrony. Given this effect, and the importance of phase coding in the hippocampus {Cite}, it is reasonable to wonder why hippocampal oscillations are GABAergic. Our model suggests that while this is true for brief, bursty excitatory oscillations, stronger, tonic excitatory oscillations engage homeostatic mechanisms that alter Na and K conductances, driving down the membrane voltage. This, in turn, reduces both the firing rate and spike synchrony in response to inputs.

@@

Bursting, rather than sustained, oscillations tend to be common in the cortex. One striking example is in motor cortical regions, where beta (~12-30 Hz) bursts are prevalent and likely functional. Specifically, beta bursts are very short--sometimes lasting only one or two cycles \cite{Sherman2016,Cole2019,Jones2016}--and relate to self-timed movements \cite{Feingold2015}. Patients with Parkinson's disease show increased motor rigidity and bradykinesia, symptoms associated with prolongation of beta bursts \cite{Tinkhauser2017}. Levodopa treatment was shown to decrease burst probability and duration, and that decrease in burst duration correlated with motor improvement \cite{Tinkhauser2017}.

An important counterexample to beta is visual cortical alpha (8-10 Hz). Alpha is tonic and high powered when humans rest with their eyes closed. Even though both cortical and sub-cortical alpha generators are synaptically excitatory, this rhythm has however been associated with suppression of excitability \cite{Jensen2002,Bonnefond2012,Peterson2017}. While several competing explanations have been offered for this \cite{Bonnefond2012,Lange2013,Peterson2017}, our work raises another possibility that is complementary to the others. Specifically, though we modeled hippocampal cells, the same Ca\textsuperscript{2+} homeostatic mechanisms exist in visual cortex (and the neocortex as a whole). This means that strong and tonic alpha oscillations, combined with homeostasis, should directly suppress population firing in visual cortex. Such an effect, were it to occur, would last well past the moment of oscillation offset. In fact, such long term effects of alpha have been reported in the literature \cite{Jensen2002,Bonnefond2012}, though the physiological mechanism was often unclear. Our work suggests that intrinsic homeostasis may underlie this effect.

@@
